---
title: "Hoefer et al. 2024 – PAM in Mammals"
author: "Sebastian Hoefer"
date: today
title-block-banner: "#161616"
title-block-banner-color: "white"
format:
  html:
    embed-resources: true
    grid: 
      body-width: 2000px
      sidebar-width: 200px
      margin-width: 300px
    toc: true
    toc-title: \deftocheading{toc}{}
    toc-depth: 6
    toc-location: right
    toc-expand: 1
    code-overflow: wrap
    code-block-bg: "F2F2F2"
    code-block-border-left: "#4A9D9E"
    highlight-style: github
    theme: darkly
    knitr:
      opts_chunk: 
        R.options:
          width: 120
standalone: true
---

# Load packages

Load packages needed for data manipulation and analyses.

```{r}
#| include: false
#| echo: true

library(tidyverse)    # Data wrangling and visualization
#library(readxl)
#library(janitor)
library(patchwork)    # Combine ggplots
library(fossil)       # Chao2 calculations
#library(sf)           # For FrogID SDM extractions
#library(galah)        # For ALA record extractions
library(mvabund)


library(colorspace)   # Color manipulation
library(cmdstanr)     # Interface to 'CmdStan'
library(brms)         # Bayesian regression models
library(rstan)        # Interface to Stan
library(ggeffects)    # Marginal effects plots
library(DHARMa)       # Residual diagnostics
library(emmeans)      # Estimated marginal means
library(tidybayes)    # Tidy data for Bayesian models
library(vegan)        # Community ecology analysis
library(EcolUtils)    # Ecology utilities
library(svglite)      # Save figures as SVG files
library(HDInterval)   # HPD intervals
library(report)       # Automated reporting
library(ggsignif)     # Significance levels in ggplot
library(ggimage)      # Images in ggplot
library(cowplot)      # Plot composition in ggplot
library(scales)       # Manipulate axis labels
```

# Functions

Custom ggplot theme for visualisation.

```{r}
my.theme <- function(){
  theme_classic() +
    theme(text = element_text(family = "Avenir Next"),
          axis.title.y = element_text(margin = margin(t = 0,r = 20,b = 0,l = 0)),
          axis.title.x = element_text(margin = margin(t = 20,r = 0,b = 0,l = 0)), 
          plot.margin = unit(c(5, 10, 5, 10), units = "mm"),
          strip.background = element_rect(fill = "#CCCCFF"),
          strip.text.x = element_text(size = 20),
          axis.title = element_text(size = 20),
          axis.text = element_text(size = 18),
          legend.text = element_text(size = 15),
          legend.title = element_text(size = 15))
}
```

# Prepare data

## Survey data

### Trip 1

```{r}
searches <- read_xlsx("../data/raw/Trip Nov-Dec 2023 tropwater data.xlsx", sheet = "searches") |>  
  mutate(across(c(start, end, time), 
                ~str_pad(sub("(\\d+)(\\d{2})", "\\1:\\2", .), 
                        width = 5, side = "left", pad = "0")))

incidentals <- read_xlsx("../data/raw/Trip Nov-Dec 2023 tropwater data.xlsx", sheet = "incidentals")  |>  
  mutate(time = str_pad(sub("(\\d+)(\\d{2})", "\\1:\\2", time), width = 5, side = "left", pad = "0"))

sp_reference <- read_csv("../data/processed/database.csv")
```

Clean up naming inconsistencies and remove unknown species IDs

```{r}
#| warning: false

#Active searches
searches_clean <- searches |> 
  filter(!common.name %in% c("Friarbird", "Fairywren", "Flying fox", "Honeyeater", "Kite", "Macropod", "Microbat", "recording 108", "Cuckooshrike", "Whipsnake")) |> 
  mutate(common.name = fct_recode(common.name,
                          "Australian Magpie" = "Australian magpie",
                          "Barking Owl" = "Barking owl",
                          "Cane Toad" = "Cane toad",
                          "Pacific Koel" = "Koel",
                          "Northern Velvet Gecko" = "Oedura castelnaui",
                          "Cape York Spotted Python" = "Spotted python",
                          "Australian Boobook" = "Southern Boobook")) |> 
  left_join(sp_reference, by = "scientific.name", suffix = c("", ".sci")) |>
  mutate(common.name = coalesce(common.name, common.name.sci)) |>
  left_join(sp_reference, by = "common.name", suffix = c("", ".com")) |> 
  mutate(scientific.name = coalesce(scientific.name, scientific.name.com),
         class = coalesce(class, class.com),
         genus = coalesce(genus, genus.com),
         family = coalesce(family, family.com)) |>
  select(-ends_with(c(".sci", ".com")))

#Incidentals
incidentals_clean <- incidentals |> 
  filter(!common.name %in% c("Monitor", "Wallaroo", "Dragon", "monitor", "Crocodiles")) |> 
  mutate(common.name = fct_recode(common.name,
                          "Emu" = "emu",
                          "Short-beaked Echidna" = "Echidna",
                          "Sahul Sunbird" = "Olive-backed Sunbird",
                          "European Rabbit" = "Rabbit",
                          "Australian Bustard" = "Australian bustard",
                          "Grey-crowned Babbler" = "Grey-crowned babbler")) |> 
  left_join(sp_reference, by = "scientific.name", suffix = c("", ".sci")) |>
  mutate(common.name = coalesce(common.name, common.name.sci)) |>
  left_join(sp_reference, by = "common.name", suffix = c("", ".com")) |> 
  mutate(scientific.name = coalesce(scientific.name, scientific.name.com),
         class = coalesce(class, class.com),
         genus = coalesce(genus, genus.com),
         family = coalesce(family, family.com)) |>
  select(-ends_with(c(".sci", ".com")))
```

Merge cleaned data frames.

```{r}
trip1 <- merge(searches_clean, incidentals_clean, all = TRUE) %>%
  add_column(season = "wet",
             trip = "trip1") |> 
  mutate(date.time = as.POSIXct(paste(date, time), format = "%Y-%m-%d %H:%M"),
    assessment.method = case_when(
    survey.type == "Birding" ~ "birding",
    survey.type == "Spotlight" ~ "spotlighting",
    TRUE ~ "incidentals"),
    site = case_when(
      plot %in% c("11","76","99","103", "107") ~ "wet_site",
      plot %in% c("23","40", "50", "70", "101") ~ "dry_site",
      TRUE ~ plot)) |> 
  uncount(number) |> 
  select(date.time, date, time, site, plot, location, detection, common.name, scientific.name, genus, family, class,assessment.method,season,trip, participants,
         temperature.C, barometric.pressure.mbar, `cloud cover.%`, max.wind.kmh, average.wind.kmh, `humidity.%`, rain, start, end, regenerated.tail, notes)
```

Save cleaned and merged data frame.

```{r}
write_csv(trip1, "../data/processed/data_trip1.csv")
```

### Trip 2

```{r}
searches <- read_xlsx("../data/raw/Trip 2 (Jul 2024).xlsx", sheet = "searches") |>  
  mutate(across(c(start, end, time), 
                ~str_pad(sub("(\\d+)(\\d{2})", "\\1:\\2", .), 
                        width = 5, side = "left", pad = "0")))

incidentals <- read_xlsx("../data/raw/Trip 2 (Jul 2024).xlsx", sheet = "incidentals")  |>  
  mutate(time = str_pad(sub("(\\d+)(\\d{2})", "\\1:\\2", time), width = 5, side = "left", pad = "0"))

sp_reference <- read_csv("../data/processed/database.csv")
```

Clean up naming inconsistencies and remove unknown species IDs

```{r}
#| warning: false

#Active searches
searches_clean <- searches |> 
  filter(!common.name %in% c("Rosella", "Microbat", "Macropod","recording 119", "recording 121","Friarbird" ,"Duck", "Honeyeater")) |> 
  mutate(common.name = str_to_title(common.name),
         common.name = str_replace_all(common.name, "-\\w+", tolower),
         common.name = fct_recode(common.name,
                          "Sahul Sunbird" = "Sunbird",
                          "Australian Boobook" = "Southern Boobook",
                          "Willie-wagtail" = "Willie Wagtail",
                          "Little Bronze-Cuckoo" = "Little Bronze Cuckoo",
                          "Green Tree Frog" = "Australian Green Treefrog",
                          "Domestic Cattle" = "Cow",
                          "Red-tailed Black-Cockatoo" = "Red-tailed Black Cockatoo")) |> 
  left_join(sp_reference, by = "scientific.name", suffix = c("", ".sci")) |>
  mutate(common.name = coalesce(common.name, common.name.sci)) |>
  left_join(sp_reference, by = "common.name", suffix = c("", ".com")) |> 
  mutate(scientific.name = coalesce(scientific.name, scientific.name.com),
         class = coalesce(class, class.com),
         genus = coalesce(genus, genus.com),
         family = coalesce(family, family.com)) |>
  select(-ends_with(c(".sci", ".com")))

#Incidentals
incidentals_clean <- incidentals |> 
  mutate(common.name = str_to_title(common.name),
         common.name = str_replace_all(common.name, "-\\w+", tolower),
         common.name = fct_recode(common.name,
                                  "Yellow-spotted Goanna" = "Yellow-spotted Monitor",
                                  "White-bellied Sea-Eagle" = "White-bellied Sea Eagle")) |> 
  left_join(sp_reference, by = "scientific.name", suffix = c("", ".sci")) |>
  mutate(common.name = coalesce(common.name, common.name.sci)) |>
  left_join(sp_reference, by = "common.name", suffix = c("", ".com")) |> 
  mutate(scientific.name = coalesce(scientific.name, scientific.name.com),
         class = coalesce(class, class.com),
         genus = coalesce(genus, genus.com),
         family = coalesce(family, family.com)) |>
  select(-ends_with(c(".sci", ".com"))) |> 
  filter(!scientific.name %in% c("Cryptoblepharus", "Demansia"))
```

Merge cleaned data frames.

```{r}
trip2 <- merge(searches_clean, incidentals_clean, all = TRUE) %>%
  add_column(season = "wet",
             trip = "trip2") |> 
  mutate(date.time = as.POSIXct(paste(date, time), format = "%Y-%m-%d %H:%M"),
    assessment.method = case_when(
    survey.type == "Birding" ~ "birding",
    survey.type == "Spotlight" ~ "spotlighting",
    TRUE ~ "incidentals"),
    site = case_when(
      plot %in% c("11","76","99","103", "107") ~ "wet_site",
      plot %in% c("23","40", "50", "70", "101") ~ "dry_site",
      TRUE ~ plot)) |> 
  uncount(number) |> 
  select(date.time, date, time, site, plot, location, detection, common.name, scientific.name, genus, family, class,assessment.method,season,trip, participants,
         temperature.C, barometric.pressure.mbar, `cloud cover.%`, max.wind.kmh, average.wind.kmh, `humidity.%`, rain, start, end, regenerated.tail, notes)
```

Save cleaned and merged data frame.

```{r}
write_csv(trip2, "../data/processed/data_trip2.csv")
```

### Combine Trips

```{r}
trip1 <- read_csv("../data/processed/data_trip1.csv")
trip2 <- read_csv("../data/processed/data_trip2.csv")

all_trips <- rbind(trip1,trip2)
```

```{r}
write_csv(all_trips, "../data/processed/data_all_trips.csv")
```

## Camera traps

Camera trapping data from Wildlife Insights

```{r}
cam <- read_csv("../data/raw/images_2007126.csv") |> 
  unite(scientific.name, c(genus, species), sep = " ", remove = FALSE) |> 
  mutate(scientific.name = case_when(
      scientific.name == "NA NA" ~ NA_character_,
    TRUE ~ scientific.name),
    scientific.name = coalesce(scientific.name, individual_animal_notes),
    scientific.name = case_when(
      family == "Felidae" ~ "Felis catus",
      common_name == "Kangaroo Family" ~ "Macropodidae",
      common_name == "Woodrat or Rat Species" ~ "Rattus",
      common_name == "Reptile" & is.na(scientific.name) ~ "Carlia munda",
      common_name == "Bandicoot" ~ "Isoodon macrourus",
      TRUE ~ scientific.name),
    scientific.name = fct_recode(scientific.name,
                          "Notamacropus agilis" = "Macropus agilis",
                          "Notamacropus parryi" = "Macropus parryi",
                          "Varanus panoptes panoptes" = "Varanus panoptes"),
    plot = case_when(deployment_id %in% c("11 2/2", "11 rest", "IB 1/2", "11 10/07/2024") ~ "11",
                     deployment_id %in% c("103 rest 0", "B6 1/2 0","103 10/07/2024") ~ "103",
                     deployment_id %in% c("107 rest 0", "Aa 1/2", "AA 2/2", "107 10/07/2024") ~ "107",
                     deployment_id %in% c("101 all", "101  10/07/2024") ~ "101",
                     deployment_id %in% c("40 07/10/2024") ~ "40",
                     deployment_id %in% c("50 rest","50 10/07/2024") ~ "50",
                     deployment_id %in% c("70 all", "70 11/24/2023") ~ "70",
                     deployment_id %in% c("23 11/26/2023", "23 10/07/2024") ~ "23",
                     deployment_id %in% c("76 all") ~ "76",
                     deployment_id %in% c("99 all", "99 10/07/2024") ~ "99"),
    site = case_when(
           plot %in% c("11","76","99","103", "107") ~ "wet_site",
           plot %in% c("23","40", "50", "70", "101") ~ "dry_site",
           TRUE ~ plot),
    season = case_when(
           month(timestamp) %in% c(10, 11, 12, 1, 2, 3, 4) ~ "wet",
           TRUE ~ "wet"),
    trip = case_when(
           timestamp >= ymd("2023-11-01") & timestamp <= ymd("2024-02-29") ~ "trip1",
           timestamp >= ymd("2024-07-01") & timestamp <= ymd("2024-09-30") ~ "trip2",
           TRUE ~ "adj_datetime"),
    date = as_date(timestamp),
    time = format(timestamp, "%H:%M:%S")) |> 
  select(date.time = timestamp, date, time, site, plot, scientific.name, class, image_id, season, trip) |> 
  add_column(detection = "seen",
             location = "on",
             assessment.method = "camera") |> 
  left_join(sp_reference)
```

```{r}
write_csv(cam, "../data/processed/camera_trapping_data.csv")
```

## PAM data

### BirdNET-Analyzer

```{r}
txt_files <- list.files(path = "../data/raw/BirdNET_detections/", pattern = "*.txt", full.names = TRUE, recursive = TRUE)

birdnet <- txt_files |>
  purrr::map(read_delim, 
      delim = "\t",  # tab as the delimiter between columns
      show_col_types = FALSE,
      col_types = cols(.default = "c")) |>  # Read all columns as character
  bind_rows() |> 
  mutate(plot = str_extract(`Begin File`, "(?<=_)[^_]+(?=_)"),
         datetime = ymd_hms(str_extract(`Begin File`, "^\\d{8}T\\d{6}"),
                            tz = "Australia/Sydney"),
         adj_datetime = datetime + seconds(`Begin Time (s)`))

```

Now we need to filter out low BirdNET Confidence scores. The threshold is extremely variable and there is no one-for all solution but Perez-Granados et al. 2023 (https://onlinelibrary.wiley.com/doi/10.1111/ibi.13193) and Sethi et al. 2021 (https://brage.nina.no/nina-xmlui/bitstream/handle/11250/2832294/ninarapport2064.pdf?sequence=4&isAllowed=y) has suggested 0.7-0.8 for most studies.

```{r}
birdnet_likely <- birdnet |> 
  filter(Confidence > 0.75)
```


BirdNET out of the box has not been trained very well with Australian bird species and therefore gets a fair few wrong. So we need to manually exclude some of those really unlikely ones based on distribution records on eBird.

BirdNET species lists for each recorder based on occurrence data and time of year. These were created with the BirdNET-Analyzer GUI and a threshold of 0.03 (3% likelihood.

```{r}
txt_files <- list.files(path = "../data/raw/BirdNet_likely_species/", pattern = "*.txt", full.names = TRUE, recursive = TRUE)

BirdNET_sp_lists <- txt_files |>
  purrr::map(read_delim, 
      delim = "_",
      col_names = c("scientific.name", "Common Name"),
      show_col_types = FALSE,
      col_types = cols(.default = "c")) |>
  bind_rows() |> 
  distinct()
```

```{r}
matching_species <- birdnet_likely |>
  semi_join(BirdNET_sp_lists, by = "Common Name") |> 
  rename(common.name = `Common Name`)
```

Even after setting the Confidence to 0.75 and restricting the species list to region and time of year, we still need to be cautious about some of the IDs and take them with a grain of salt. They would likely need validation to be sure as the Australian birds in BirdNET aren't as well represented and trained for.

Now we need to clean up the species names.

```{r}
sp_reference <- read_csv("../data/processed/database.csv")

birdnet_cleaned <- matching_species |> 
  mutate(common.name = str_replace(common.name, "Gray", "Grey"),
         common.name = fct_recode(common.name,
                                  "Australian Boobook" = "Southern Boobook",
                                  "Bush Stone-curlew" = "Bush Thick-knee",
                                  "Beach Stone-curlew" = "Beach Thick-knee",
                                  "Sahul Sunbird" = "Olive-backed Sunbird",
                                  "Eastern Barn Owl" = "Barn Owl",
                                  "Sahul Brush Cuckoo" = "Brush Cuckoo",
                                  "Sahul Cicadabird" = "Common Cicadabird",
                                  "Papuan Eclectus" = "Eclectus Parrot",
                                  "Australian Rufous Fantail" = "Rufous Fantail",
                                  "Eastern Cattle-Egret" = "Cattle Egret",
                                  "Australian Spectacled Monarch" = "Spectacled Monarch")) |> 
  left_join(sp_reference, by = "common.name") |> 
  mutate(plot = str_remove(plot, "^R"),
         site = case_when(
           plot %in% c("11","76","99","103", "107") ~ "wet_site",
           plot %in% c("23","40", "50", "70", "101") ~ "dry_site",
           TRUE ~ plot),
         season = case_when(
           month(adj_datetime) %in% c(10, 11, 12, 1, 2, 3, 4) ~ "wet",
           TRUE ~ "wet"),
         trip = case_when(
           adj_datetime >= ymd("2023-11-01") & adj_datetime <= ymd("2024-02-29") ~ "trip1",
           adj_datetime >= ymd("2024-07-01") & adj_datetime <= ymd("2024-09-30") ~ "trip2",
           TRUE ~ "adj_datetime"),
         `Begin Time (s)` = as.numeric(`Begin Time (s)`),
         `End Time (s)` = as.numeric(`End Time (s)`),
         file_number = str_extract(`Begin File`, "\\d+(?=\\.wav)"),
         AudioLink = sprintf("https://api.ecosounds.org/audio_recordings/%s/media.wav?start_offset=%.1f&end_offset=%.1f",
                             file_number,
                             `Begin Time (s)` - 3.5,
                             `End Time (s)` + 3.5),
         file_number = NULL) |> 
  select(date.time = adj_datetime, site, plot,common.name, scientific.name,genus, family, trip, season, Confidence, AudioLink)
```

```{r}
write_csv(birdnet_cleaned, "../data/processed/pam_data.csv")
```

### Embeddings

#### FrogID species prediction

Extract a frog species list based on FrogID SDMs and the recording locations to know what frogs to look for.

```{r}

# Get list of .shp files
shp_files <- list.files(path = "/Users/sebastianhoefer/Documents/Research/GIS/FrogID_species_distribution/2024/",
                        pattern = "\\.shp$", full.names = TRUE)

# Read in study_sites data
sites <- read_csv("../data/processed/study_sites.csv")

survey_sites_sf <- st_as_sf(sites, coords = c("lon", "lat"), crs=4326)


df_list <- map(shp_files, function(file) {
  tryCatch(
    {
      # Read .shp file
      t <- st_read(file)
      
      # Check and try to fix if invalid
      if (any(!st_is_valid(t))) {
        message(paste("Invalid geometries in", file))
        
        if ("lwgeom" %in% installed.packages()) {
          library(lwgeom)
          t <- st_make_valid(t)
        } else {
          t <- t[st_is_valid(t), ]
        }
      }
      
      # Transform 
      t <- st_transform(t, 4326)
      
      # Operation
      joined_data <- st_join(survey_sites_sf, t, left=TRUE)
      
      # Get the file name without the extension
      filename <- tools::file_path_sans_ext(basename(file))
      
      # Check if the Species column exists
      if ("Species" %in% names(joined_data)) {
        joined_data2 <- joined_data %>%
          mutate(presence = if_else(is.na(Species), "0", "1")) %>%
          mutate(Species = filename) %>%
          select(location, Species, presence)
        
        # print some metrics for assessment
        message(paste("Processed ", file, " with ", nrow(joined_data2), " rows and ", length(names(joined_data2)), " columns."))
        
        return(joined_data2)
      } else {
        message(paste("Species column not found in", file))
        return(tibble())  # return an empty dataframe
      }
    },
    error = function(cond) {
      message(paste("Error in processing file: ", file))
      message("Here is the original error message:")
      message(cond$message)
      return(tibble())  # Returns an empty dataframe, and still passes to the next file
    }
  )
})


frogs <- bind_rows(df_list) %>% 
  mutate(Species = str_replace_all(Species, "_", " "),
         presence = as.numeric(presence)) %>% 
  st_set_geometry(NULL) %>% 
  pivot_wider(names_from = "location", values_from = "presence") %>% 
  replace(is.na(.), "0") %>% 
  select(Species, everything()) %>%
  filter(rowSums(select(.,where(is.numeric))) > 0)
  
write_csv(frogs, "../data/processed/FrogID_SDM_presence.csv")
```

#### Data preparation

```{r}

```


# ANALYSIS

## Load data

```{r}
tropwater_data <- read_csv("../data/processed/data_all_trips.csv") |> 
  filter(!plot %in% c("Road", "Einasleigh")) |> 
  mutate(plot = as.factor(plot))

pam_data <- read_csv("../data/processed/pam_data.csv") |> 
  add_column(assessment.method = "PAM",
             class = "Aves") |> 
  mutate(date = as_date(date.time),
    time = hms::as_hms(date.time),
    plot = as.factor(plot))

cam_data <- read_csv("../data/processed/camera_trapping_data.csv") |> 
  mutate(plot = as.factor(plot)) |> 
  filter(!scientific.name %in% c("Rattus", "Macropodidae", "Coturnix NA"))

#Join all data frames

data <- list(tropwater_data, cam_data, pam_data) |> 
  reduce(full_join) |> 
  mutate(survey.method = assessment.method,
         assessment.method = case_when(
           assessment.method == "PAM" ~ "PAM",
           assessment.method == "camera" ~ "camera",
           assessment.method %in% c("birding", "spotlighting", "incidentals") ~ "OBM"))
```

## Abundance

```{r}
#| fig-width: 15
#| fig-asp: 0.4

(abund_plot <- ggplot(data, aes(class, fill = site)) +
  geom_bar(position = position_dodge2(width = 0.6)) +
  facet_wrap(~assessment.method, scales = "free_y") +
  my.theme() +
  scale_y_continuous(name = "Detections",
                     labels = scales::comma) +
  scale_x_discrete(name = "")) +
  scale_fill_manual(values = c("coral", "steelblue")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.spacing = unit(2, "lines"))
```

```{r}
#| fig-width: 15
#| fig-asp: 0.6

(abund_plot2 <- ggplot(data, aes(trip, fill = site)) +
  geom_bar(position = position_dodge2(width = 0.6)) +
  facet_wrap(~class, scales = "free_y") +
  my.theme() +
  scale_y_continuous(name = "Detections",
                     labels = scales::comma) +
  scale_x_discrete(name = "")) +
  scale_fill_manual(values = c("coral", "steelblue")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.spacing = unit(2, "lines"))
```

```{r}
ggsave("../output/abund_plot_trips.png", abund_plot2, width = 15, height = 15*0.6, dpi = 300)
```

## Species richness

```{r}
sp_rich <- data |> 
  group_by(site, common.name, class, assessment.method) |>  
  distinct(scientific.name) |>  
  ungroup()

sp_rich_b <- data |> 
  group_by(site, common.name, class, trip) |>  
  distinct(scientific.name) |>  
  ungroup()
```

```{r}
#| fig-width: 15
#| fig-asp: 0.4

(richness_plot <- ggplot(sp_rich, aes(class, fill = site)) +
  geom_bar(position = position_dodge2(width = 0.6)) +
  facet_wrap(~assessment.method, scales = "free_y") +
  my.theme() +
  scale_y_continuous(name = "Species Richness",
                     labels = scales::comma) +
  scale_x_discrete(name = "")) +
  scale_fill_manual(values = c("coral", "steelblue")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.spacing = unit(2, "lines"))
```

```{r}
#| fig-width: 15
#| fig-asp: 0.6

(richness_plot2 <- ggplot(sp_rich_b, aes(trip, fill = site)) +
  geom_bar(position = position_dodge2(width = 0.6)) +
  facet_wrap(~class, scales = "free_y") +
  my.theme() +
  scale_y_continuous(name = "Species Richness",
                     labels = scales::comma) +
  scale_x_discrete(name = "")) +
  scale_fill_manual(values = c("coral", "steelblue")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.spacing = unit(2, "lines"))
```

```{r}
ggsave("../output/richness_plot_trips.png", richness_plot2, width = 15, height = 15*0.6, dpi = 300)
```

## Combined plots

```{r}
#| fig-width: 20
#| fig-asp: 0.8

(combined_plots <- richness_plot / abund_plot +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom",
        legend.box = "horizontal",
        axis.text.x = element_text(angle = 45, hjust = 1)) &
  scale_fill_manual(values = c("coral", "steelblue")))
```

```{r}
ggsave("../output/richness_plots.png", combined_plots, width = 20, height = 20*0.8, dpi = 300)
```

```{r}
sp_rich2 <- tropwater_data |> 
  group_by(season, common.name, class) |> 
  summarise(richness = length(unique(scientific.name)))
```


## Chao2 species richness

```{r}
tropwater_data <- read_csv("../data/processed/data_all_trips.csv")

data_summed <- tropwater_data |> 
  filter(!plot %in% c("Road", "Einasleigh")) |> 
  group_by(season, scientific.name, class) |> 
  count(common.name) |> 
  ungroup()
```

```{r}
presence_absence <- data_summed %>%
  # Select the relevant columns
  select(common.name, season, class) %>%
  # Mark presence (1) for each species per site, plot and date
  distinct() %>%
  mutate(Presence = 1) %>%
  # Reshape the data into wide format
  pivot_wider(names_from = common.name, 
              values_from = Presence, 
              values_fill = list(Presence = 0))
```



Number of recording days

```{r}
pam_data |> 
  group_by(season, plot) |> 
  summarise(unique_days = n_distinct(as_date(adj_datetime)))
```

## TRYING STUFF

```{r}
tropwater_data <- read_csv("../data/processed/data_all_trips.csv")

data_summed <- tropwater_data |> 
  filter(!plot %in% c("Road", "Einasleigh")) |> 
  group_by(date, site, plot,trip, season, scientific.name) |> 
  count(common.name) |> 
  ungroup()
```

```{r}
presence_absence <- data_summed %>%
  # Select the relevant columns
  select(common.name, site, plot, date) %>%
  # Mark presence (1) for each species per site, plot and date
  distinct() %>%
  mutate(Presence = 1) %>%
  # Reshape the data into wide format
  pivot_wider(names_from = common.name, 
              values_from = Presence, 
              values_fill = list(Presence = 0))
```


```{r}
# Create function to process each site-plot combination
process_sac <- function(data, site_name, plot_name) {
  # Skip if no species present
  if (ncol(data) == 0 || nrow(data) == 0) return(NULL)
  
  data_matrix <- data |>
    mutate(across(everything(), as.numeric)) |>
    as.matrix()
  
  # Calculate SAC and Chao2
  sac <- specaccum(data_matrix, method = "random", permutations = 100)
  chao2_est <- chao2(data_matrix, taxa.row = FALSE)
  
  # Create plot data
  sac_df <- tibble(
    Richness = sac$richness,
    Sites = seq_along(sac$richness),
    SD = sac$sd,
    site = site_name,
    plot = plot_name
  )
  
  # Create and save plot
  p <- ggplot(sac_df, aes(x = Sites, y = Richness)) +
    geom_line() +
    geom_point() +
    geom_errorbar(aes(ymin = Richness - SD, ymax = Richness + SD), width = 0.2) +
    labs(title = paste("SAC for", site_name, "-", plot_name),
         x = "Number of Sites",
         y = "Species Richness") +
    my.theme()
  
  # Save outputs
  write_csv(data, 
           file = str_glue("output/presence_absence_{str_replace_all(site_name, ' ', '_')}_{str_replace_all(plot_name, ' ', '_')}.csv"))
  
  ggsave(str_glue("output/SAC_{str_replace_all(site_name, ' ', '_')}_{str_replace_all(plot_name, ' ', '_')}.png"),
         plot = p,
         width = 8, height = 5, dpi = 300)
  
  # Return results
  list(
    sac = sac,
    chao2 = chao2_est,
    plot = p,
    data = sac_df
  )
}

# Then run the full process
results <- presence_absence |>
  group_by(site, plot) |>
  nest() |>
  mutate(
    results = purrr::pmap(
      .l = list(
        data = data,
        site_name = site,
        plot_name = plot
      ),
      .f = process_sac
    )
  ) |>
  ungroup()

# Extract results
survey_sac_list <- results |>
  filter(!map_lgl(results, is.null)) |>
  pull(results) |>
  purrr::map("sac")

survey_chao2_list <- results |>
  filter(!map_lgl(results, is.null)) |>
  pull(results) |>
  purrr::map("chao2")
```

```{r}
process_sac <- function(data, site_name, plot_name) {
  print(paste("Processing:", site_name, plot_name))
  print("Data structure:")
  print(str(data))
  
  # Skip if no species present
  if (ncol(data) == 0 || nrow(data) == 0) {
    print("Skipping: no data")
    return(NULL)
  }
  
  # Convert data to numeric matrix
  data_matrix <- data |>
    mutate(across(everything(), as.numeric)) |>
    as.matrix()
  
  print("Converted data structure:")
  print(str(data_matrix))
  
  tryCatch({
    print("Attempting SAC calculation...")
    sac <- specaccum(data_matrix, permutations = 100)
    print("SAC calculated successfully")
    
    print("Attempting Chao2 calculation...")
    chao2_est <- chao2(data_matrix, taxa.row = FALSE)
    print("Chao2 calculated successfully")
    
    # Create plot data
    sac_df <- tibble(
      Richness = sac$richness,
      Sites = seq_along(sac$richness),
      SD = sac$sd,
      site = site_name,
      plot = plot_name
    )
    
    # Create and save plot
    p <- ggplot(sac_df, aes(x = Sites, y = Richness)) +
      geom_line() +
      geom_point() +
      geom_errorbar(aes(ymin = Richness - SD, ymax = Richness + SD), width = 0.2) +
      labs(title = paste("SAC for", site_name, "-", plot_name),
           x = "Number of Sites",
           y = "Species Richness") +
      theme_minimal()
    
    list(
      sac = sac,
      chao2 = chao2_est,
      plot = p,
      data = sac_df
    )
  }, error = function(e) {
    print(paste("Error occurred:", e$message))
    return(NULL)
  })
}

# Then run the full process
results <- presence_absence |>
  group_by(site, plot) |>
  nest() |>
  mutate(
    results = purrr::pmap(
      .l = list(
        data = data,
        site_name = site,
        plot_name = plot
      ),
      .f = process_sac
    )
  ) |>
  ungroup()

# Extract results
survey_sac_list <- results |>
  filter(!map_lgl(results, is.null)) |>
  pull(results) |>
  purrr::map("sac")

survey_chao2_list <- results |>
  filter(!map_lgl(results, is.null)) |>
  pull(results) |>
  purrr::map("chao2")

```

```{r}
(sac_plots <- results |>
  filter(!map_lgl(results, is.null)) |>
  pull(results) |>
  purrr::map("plot"))
```

```{r}
sac_data <- results |>
  filter(!map_lgl(results, is.null)) |>
  pull(results) |>
  map_dfr("data", .id = "group")
```


```{r}
unique_locations <- unique(presence_absence$Location)

# Create a list to store the SACs
survey_sac_list <- list()
survey_chao2_list <- list()

# Loop through each unique Location
for (location in unique_locations) {
  
  # Filter data for the current Location
  location_data <- presence_absence %>%
    filter(Location == location)
  
  # Get unique Points within this Location
  unique_points <- unique(location_data$Point)
  
  # Loop through each Point within the current Location
  for (point in unique_points) {
    
    # Filter data for this specific Location and Point combination
    point_data <- location_data %>%
      filter(Point == !!point) %>%
      select(-Location, -Point, -date)  # Drop non-species columns
    
    # Skip if there are no species present
    if (ncol(point_data) == 0 || nrow(point_data) == 0) {
      next
    }
    
    # Save the presence-absence matrix as a CSV file
    write_csv(point_data, 
              file = paste0("presence_absence_", gsub(" ", "_", location), "_", gsub(" ", "_", point), ".csv"))
    
    # Calculate species accumulation curve
    sac <- specaccum(point_data, permutations = 100)  # Increase permutations for a better estimate
    
    # Store the result in the list
    survey_sac_list[[paste(location, point, sep = "_")]] <- sac
    
    # Calculate Chao2 richness estimate
    chao2_estimate <- chao2(point_data, taxa.row = FALSE)
    survey_chao2_list[[paste(location, point, sep = "_")]] <- chao2_estimate
    
    # Create a data frame for ggplot
    sac_df <- data.frame(
      Richness = sac$richness,
      Sites = 1:length(sac$richness),
      SD = sac$sd  # Standard deviations
    )
    
    # Generate the plot
    p <- ggplot(sac_df, aes(x = Sites, y = Richness)) +
      geom_line() +
      geom_point() +
      geom_errorbar(aes(ymin = Richness - SD, ymax = Richness + SD), width = 0.2) +  # Error bars
      labs(title = paste("SAC for", location, "-", point),
           x = "Number of Sites",
           y = "Species Richness") +
      theme_minimal()
    
    # Print the plot (optional)
    print(p)
    
    # Save the plot as a PNG file (optional)
    ggsave(filename = paste0("SAC_", gsub(" ", "_", location), "_", gsub(" ", "_", point), ".png"),
           plot = p,
           width = 8, height = 5, dpi = 300)
  }
}

survey_sac_list

# Initialize an empty list to store the data frames
survey_sac_data_list <- list()

# Loop through the sac_list to extract relevant data
for (plot in names(survey_sac_list)) {
  sac <- survey_sac_list[[plot]]
  
  # Create a data frame for the current plot
  sac_df <- data.frame(
    Plot = plot,
    Sites = 1:length(sac$richness),
    Richness = sac$richness,
    SD = sac$sd
  )
  
  # Append to the list
  survey_sac_data_list[[plot]] <- sac_df
}

# Combine all data frames into one
survey_sac_combined_df <- bind_rows(survey_sac_data_list)

# View the combined data frame
print(survey_sac_combined_df)

# Optionally, save the combined data frame to a CSV file
write.csv(sac_combined_df, "BirdSurvey_sac_combined_data.csv", row.names = FALSE)


survey_chao2_list


# Convert chao2_list to a data frame
chao2_df <- do.call(rbind, lapply(names(survey_chao2_list), function(name) {
  # Extract location and point from the name
  location_point <- strsplit(name, "_")[[1]]
  
  # Create a data frame for each entry
  data.frame(
    Location = location_point[1],
    Point = location_point[2],
    Chao2_Estimate = survey_chao2_list[[name]],
    stringsAsFactors = FALSE
  )
}))

# View the resulting data frame
print(chao2_df)

# Optionally save the data frame as a CSV file
write_csv(chao2_df, "BirdSurvey_chao2_estimates.csv")
```

## Ordination

### NMDS

```{r}

```

### MVABUND

#### BioClim

Extract BioClim data for MVAbund analysis.

Load packages

```{r}
library(geodata)
library(terra)
library(sp)
```

```{r}
# Download WorldClim data (BioClim variables at 10 arc minutes resolution)
worldclim_data <- worldclim_global(var = "bio", res = 10, path = tempdir())

# Define the coordinates (longitude, latitude) of interest
coords <- read_csv("../data/processed/study_sites.csv")

#convert to a geo object
survey_sites_sf <- st_as_sf(sites, coords = c("lon", "lat"), crs=4326)

# Extract the WorldClim data for these points
extracted_data <- extract(worldclim_data, survey_sites_sf)

# Combine the coordinates with the extracted data
result <- cbind(coords, extracted_data)

# Export the dataframe as a CSV file
write_csv(result, "../data/processed/site_clim_data.csv")
```

Meaning of Bioclim variables:

BIO1 = Annual Mean Temperature

BIO2 = Mean Diurnal Range (Mean of monthly (max temp - min temp))

BIO3 = Isothermality (BIO2/BIO7) (×100)

BIO4 = Temperature Seasonality (standard deviation ×100)

BIO5 = Max Temperature of Warmest Month

BIO6 = Min Temperature of Coldest Month

BIO7 = Temperature Annual Range (BIO5-BIO6)

BIO8 = Mean Temperature of Wettest Quarter

BIO9 = Mean Temperature of Driest Quarter

BIO10 = Mean Temperature of Warmest Quarter

BIO11 = Mean Temperature of Coldest Quarter

BIO12 = Annual Precipitation

BIO13 = Precipitation of Wettest Month

BIO14 = Precipitation of Driest Month

BIO15 = Precipitation Seasonality (Coefficient of Variation)

BIO16 = Precipitation of Wettest Quarter

BIO17 = Precipitation of Driest Quarter

BIO18 = Precipitation of Warmest Quarter

BIO19 = Precipitation of Coldest Quarter

#### Analysis

```{r}

```

